---
title: "Kruskal_wallis"
format: html
---

## Library and Data

```{r}
#| label: install_packages

#install.packages("tidyverse")
#install.packages("rstatix")
#install.packages("ggpubr")
```

```{r}
#| label: load_packages

library(tidyverse)
library(rstatix)
library(ggpubr)
library(knitr)
library(readxl)
library(data.table)
library(vegan)
library(car)
```

```{r}
#| label: read_csvs

Corals_clean <- read.csv("/Users/ellaandonov/World_Corals/Cleaned data CSVs/Corals_clean.csv")
ITS2_clean   <- read.csv("/Users/ellaandonov/World_Corals/Cleaned data CSVs/ITS2_clean.csv")
met_summary_coral_only <- read.csv("/Users/ellaandonov/World_Corals/Metabolite Summary Data/met_summary_coral_only.csv")
```

## **Kruskal wallis test:**

### Richness

```{r}
#| label: set_up_richness_metric

#Identify metabolite columns (start with "x")
metabolite_cols <- grep("^x", names(Corals_clean), value = TRUE)

# Compute metabolomic richness per sample
# Richness = count of metabolites with intensity > 0
kw_richness <- Corals_clean %>%
  rowwise() %>%
  mutate(MetabolomicRichness = sum(c_across(all_of(metabolite_cols)) > 0, na.rm = TRUE)) %>%
  ungroup()

#brought richness column to front for vizualization 
kw_richness <- kw_richness %>% select(MetabolomicRichness, everything())

```

```{r}
#| label: kw_test_richness_by_phylum

kruskal.test(host_phylum ~ MetabolomicRichness, data = kw_richness)
```

```{r}
#| label: kw_test_richness_by_location

kruskal.test(location ~ MetabolomicRichness, data = kw_richness)
```

```{r}
#| label: kw_test_richness_by_bleaching

kruskal.test(bleaching ~ MetabolomicRichness, data = kw_richness)
```

```{r}
#| label: kw_test_richness_by_symbiont_potential

kruskal.test(symbiont.potential ~ MetabolomicRichness, data = kw_richness)
```

-   **Results:** No significant difference was found in the metabolomic richness (number of metabolites present) between different phyla, locations, corals with different bleaching status, or their symbiont.

-   **Null hypothesis (H₀):** The number of metabolites are not significantly different across metadata groups, meaning the population medians are not significantly different.

    -   p-value \> 0.05, failed to reject H₀ → no evidence that groups differ.

### Entropy

```{r}
#| label: entropy_formula_setup

shannon_index <- function(counts, assume_proportions = FALSE, log_base = exp(1)) {
  # remove NAs and non-positive values
  counts <- counts[!is.na(counts) & counts > 0]
  if (length(counts) == 0) return(0)

  if (!assume_proportions) {
    total <- sum(counts)
    if (total == 0) return(0)
    p <- counts / total
  } else {
    p <- counts / sum(counts)  # if already proportions, this will still normalize safely
  }

  -sum(p * log(p, base = log_base))
  }

# identify metabolite columns (start with "x")
metabolite_cols <- grep("^x", names(Corals_clean), value = TRUE)

# calculate metabolic entropy per sample
kw_entropy <- Corals_clean %>%
  rowwise() %>%
  mutate(MetabolicEntropy = shannon_index(c_across(all_of(metabolite_cols)))) %>%
  ungroup()

#brought richness column to front for vizualization 
kw_entropy <- kw_entropy %>% select(MetabolicEntropy, everything())
```

```{r}
#| label: kw_test_entropy_by_phylum

kruskal.test(host_phylum ~ MetabolicEntropy, data = kw_entropy)
```

```{r}
#| label: kw_test_entropy_by_location

kruskal.test(location ~ MetabolicEntropy, data = kw_entropy)
```

```{r}
#| label: kw_test_entropy_by_bleaching

kruskal.test(bleaching ~ MetabolicEntropy, data = kw_entropy)
```

```{r}
#| label: kw_test_entropy_by_symbiont_potential

kruskal.test(symbiont.potential ~ MetabolicEntropy, data = kw_entropy)
```

-   **Results:** No significant difference was found in the metabolomic entropy between different phyla, locations, corals with different bleaching status, or their symbiont.

-   **Null hypothesis (H₀):** The diversity of metabolites are not significantly different across metadata groups, meaning the population medians are not significantly different.

    -   p-value \> 0.05, failed to reject H₀ → no evidence that groups differ.

### Evenness

```{r}
#| label: evenness_formula_setup

shannon_index <- function(counts, assume_proportions = FALSE, log_base = exp(1)) {
  # remove NAs and non-positive values
  counts <- counts[!is.na(counts) & counts > 0]
  if (length(counts) == 0) return(0)

  if (!assume_proportions) {
    total <- sum(counts)
    if (total == 0) return(0)
    p <- counts / total
  } else {
    p <- counts / sum(counts)  # if already proportions, this will still normalize safely
  }

  -sum(p * log(p, base = log_base))
  }

# 1) Identify metabolite columns
metabolite_cols <- grep("^x", names(Corals_clean), value = TRUE)

# 2) Compute Shannon, Richness, and Evenness per sample
kw_evenness <- Corals_clean %>%
  rowwise() %>%
  mutate(
    Richness = sum(c_across(all_of(metabolite_cols)) > 0, na.rm = TRUE),
    Shannon  = shannon_index(c_across(all_of(metabolite_cols))),
    Evenness = ifelse(Richness > 1, Shannon / log(Richness), 0)
  ) %>%
  ungroup()


#brought richness column to front for vizualization 
kw_evenness <- kw_evenness %>% select(Evenness, everything())
```

```{r}
#| label: kw_test_evenness_by_phylum

kruskal.test(host_phylum ~ Evenness, data = kw_evenness)
```

```{r}
#| label: kw_test_evenness_by_location

kruskal.test(location ~ Evenness, data = kw_evenness)
```

```{r}
#| label: kw_test_evenness_by_bleaching

kruskal.test(bleaching ~ Evenness, data = kw_evenness)
```

```{r}
#| label: kw_test_evenness_by_symbiont_potential

kruskal.test(symbiont.potential ~ Evenness, data = kw_evenness)
```

## **Bar graphs to** visualize normality

### Richness

```{r}
#| label: bar-graph-symbiont-potential 

kw_richness |>
  drop_na(symbiont.potential) |>
  ggplot(aes(x = MetabolomicRichness, fill = symbiont.potential)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(
    x= "Metabolite Richness", 
    y = "Count", 
    title = "Metabolite Richness by Symbiont Potential", 
    fill = "Symbiont Potential" 
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: bar-graph-phylum

kw_richness |>
  drop_na(host_phylum) |>
  ggplot(aes(x = MetabolomicRichness, fill = host_phylum)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(
    x= "Metabolite Richness", 
    y = "Count", 
    title = "Metabolite Richness by Phylum", 
    fill = "Phylum" 
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: bar-graph-bleaching

kw_richness |>
  drop_na(bleaching) |>
  ggplot(aes(x = MetabolomicRichness, fill = bleaching)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(
    x= "Metabolite Richness", 
    y = "Count", 
    title = "Metabolite Richness by Bleaching Status", 
    fill = "Bleaching Status" 
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: bar-graph-location

kw_richness |>
  drop_na(location) |>
  ggplot(aes(x = MetabolomicRichness, fill = location)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(
    x= "Metabolite Richness", 
    y = "Count", 
    title = "Metabolite Richness by Location", 
    fill = "Location" 
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: bar-graph-Scleractinia-symbiont-potential

kw_richness |>
  filter(host_order == "Scleractinia") |>
  drop_na(symbiont.potential) |>
  ggplot(aes(x = MetabolomicRichness, fill = symbiont.potential)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(
    x= "Metabolite Richness", 
    y = "Count", 
    title = "Metabolite Richness (Scleractinia only) by Symbiont Potential", 
    fill = "Symbiont Potential" 
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: bar-graph-Scleractinia-bleaching

kw_richness |>
  filter(host_order == "Scleractinia") |>
  drop_na(bleaching) |>
  ggplot(aes(x = MetabolomicRichness, fill = bleaching)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(
    x= "Metabolite Richness", 
    y = "Count", 
    title = "Metabolite Richness (Scleractinia only) by Bleaching", 
    fill = "Bleaching" 
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: bar-graph-Scleractinia-location

kw_richness |>
  filter(host_order == "Scleractinia") |>
  drop_na(symbiont.potential) |>
  ggplot(aes(x = MetabolomicRichness, fill = location)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(
    x= "Metabolite Richness", 
    y = "Count", 
    title = "Metabolite Richness (Scleractinia only) by Location", 
    fill = "Location" 
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

## **QQ plots to visualize normality** 

### Richness

```{r}
#| label: qq_plot_richness_symbiont_potential

kw_richness |>
  drop_na(symbiont.potential) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~symbiont.potential) +
  labs(
    title = "QQ Plots by Symbiont Potential", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
#| label: qq_plot_richness_phylum 

kw_richness |>
  drop_na(host_phylum) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~host_phylum) +
  labs(
    title = "QQ Plots by Phylum", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
#| label: qq_plot_richness_location

kw_richness |>
  drop_na(location) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~location) +
  labs(
    title = "QQ Plots by Location", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_richness_bleaching

kw_richness |>
  drop_na(bleaching) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~bleaching) +
  labs(
    title = "QQ Plots by Bleaching", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_richness_Scleractinia_family

kw_richness |>
  filter(host_order == "Scleractinia") |>
  drop_na(host_family) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~host_family) +
  labs(
    title = "QQ Plots (Scleractinia only) by Family", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_richness_Scleractinia_symbiont_potential

kw_richness |>
  filter(host_order == "Scleractinia") |>
  drop_na(symbiont.potential) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~symbiont.potential) +
  labs(
    title = "QQ Plots (Scleractinia only) by Symbiont Potential", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_richness_Scleractinia_bleaching

kw_richness |>
  filter(host_order == "Scleractinia") |>
  drop_na(bleaching) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~bleaching) +
  labs(
    title = "QQ Plots (Scleractinia only) by Bleaching", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_richness_Scleractinia_location

kw_richness |>
  filter(host_order == "Scleractinia") |>
  drop_na(location) |>
  ggplot(aes(sample = MetabolomicRichness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~location) +
  labs(
    title = "QQ Plots (Scleractinia only) by Location", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Metabolite Richness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Evenness

```{r}
#| label: qq_plot_evenness_symbiont_potential

kw_evenness |>
  drop_na(symbiont.potential) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~symbiont.potential) +
  labs(
    title = "QQ Plots by Symbiont Potential", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_evenness_bleaching

kw_evenness |>
  drop_na(bleaching) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~bleaching) +
  labs(
    title = "QQ Plots by Bleaching", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_evenness_location

kw_evenness |>
  drop_na(location) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~location) +
  labs(
    title = "QQ Plots by Location", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_evenness_phylum

kw_evenness |>
  drop_na(host_phylum) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~host_phylum) +
  labs(
    title = "QQ Plots by Phylum", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_evenness_Scleractinia_family

kw_evenness |>
  filter(host_order == "Scleractinia") |>
  drop_na(host_family) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~host_family) +
  labs(
    title = "QQ Plots (Scleractinia only) by Family", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_evenness_Scleractinia_symbiont_potential 

kw_evenness |>
  filter(host_order == "Scleractinia") |>
  drop_na(symbiont.potential) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~symbiont.potential) +
  labs(
    title = "QQ Plots (Scleractinia only) by Symbiont Potential", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_evenness_Scleractinia_bleaching

kw_evenness |>
  filter(host_order == "Scleractinia") |>
  drop_na(bleaching) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~bleaching) +
  labs(
    title = "QQ Plots (Scleractinia only) by Bleaching", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_evenness_Scleractinia_location

kw_evenness |>
  filter(host_order == "Scleractinia") |>
  drop_na(location) |>
  ggplot(aes(sample = Evenness)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~location) +
  labs(
    title = "QQ Plots (Scleractinia only) by Location", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Evenness"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Entropy

```{r}
#| label: qq_plot_entropy_phylum

kw_entropy |>
  drop_na(host_phylum) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~host_phylum) +
  labs(
    title = "QQ Plots by Phylum", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_entropy_symbiont_potential

kw_entropy |>
  drop_na(symbiont.potential) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~symbiont.potential) +
  labs(
    title = "QQ Plots by Symbiont Potential", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_entropy_bleaching

kw_entropy |>
  drop_na(bleaching) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~bleaching) +
  labs(
    title = "QQ Plots by Bleaching", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_entropy_location

kw_entropy |>
  drop_na(location) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~location) +
  labs(
    title = "QQ Plots by Location", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_entropy_Scleractinia_family

kw_entropy |>
  filter(host_order == "Scleractinia") |>
  drop_na(host_family) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~host_family) +
  labs(
    title = "QQ Plots (Scleractinia only) by Family", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_entropy_Scleractinia_symbiont_potential

kw_entropy |>
  filter(host_order == "Scleractinia") |>
  drop_na(symbiont.potential) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~symbiont.potential) +
  labs(
    title = "QQ Plots (Scleractinia only) by Symbiont Potential", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_entropy_Scleractinia_bleaching

kw_entropy |>
  filter(host_order == "Scleractinia") |>
  drop_na(bleaching) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~bleaching) +
  labs(
    title = "QQ Plots (Scleractinia only) by Bleaching", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: qq_plot_entropy_Scleractinia_location

kw_entropy |>
  filter(host_order == "Scleractinia") |>
  drop_na(location) |>
  ggplot(aes(sample = MetabolicEntropy)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~location) +
  labs(
    title = "QQ Plots (Scleractinia only) by Location", 
    x = "Theoretical Quantiles (Normal Distribution)",
    y = "Entropy"
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Shaprio Wilks **for normality** 

### Richness

```{r}
#| label: shaprio_wilks_richness_phylum

kw_richness |>
  group_by(host_phylum) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )

```

```{r}
#| label: shaprio_wilks_richness_symbiont_potential

kw_richness |>
  group_by(symbiont.potential) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_richness_location

kw_richness |>
  group_by(location) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_richness_bleaching

kw_richness |>
  filter(bleaching != "Pale") |>
  group_by(bleaching) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_richness_Scleractinia_symbiont_potential

kw_richness |>
  filter(symbiont.potential != "Aposymbiotic") |>
  filter(host_order == "Scleractinia") |>
  group_by(symbiont.potential) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_richness_Scleractinia_bleaching

kw_richness |>
  filter(!is.na(bleaching), host_order == "Scleractinia") |>
  group_by(bleaching) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_richness_Scleractinia_location

kw_richness |>
  filter(host_order == "Scleractinia") |>
  group_by(location) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_richness_Scleractinia_family

kw_richness |>
  filter(host_family != "Plerogyridae", host_order == "Scleractinia") |>
  group_by(host_family) |>
  summarise(
    W = shapiro.test(MetabolomicRichness)$statistic,
    p_value = shapiro.test(MetabolomicRichness)$p.value
  )
```

### Evenness

```{r}
#| label: shaprio_wilks_evenness_phylum

kw_evenness |>
  group_by(host_phylum) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_evenness_symbiont_potential

kw_evenness |>
  group_by(symbiont.potential) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_evenness_bleaching

kw_evenness |>
  filter(bleaching != "Pale") |>
  group_by(bleaching) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_evenness_location

kw_evenness |>
  group_by(location) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_evenness_Scleractinia_family

kw_evenness |>
  filter(host_family != "Plerogyridae", host_order == "Scleractinia") |>
  group_by(host_family) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_evenness_Scleractinia_symbiont_potential

kw_evenness |>
  filter(symbiont.potential != "Aposymbiotic", host_order == "Scleractinia") |>
  group_by(symbiont.potential) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_evenness_Scleractinia_bleaching

kw_evenness |>
  filter(!is.na(bleaching), host_order == "Scleractinia") |>
  group_by(bleaching) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

```{r}
#| label: shaprio_wilks_evenness_Scleractinia_location

kw_evenness |>
  filter(host_order == "Scleractinia") |>
  group_by(location) |>
  summarise(
    W = shapiro.test(Evenness)$statistic,
    p_value = shapiro.test(Evenness)$p.value
  )
```

### Entropy

```{r}
#| label: shaprio_wilks_entropy_phylum

kw_entropy |>
  group_by(host_phylum) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

```{r}
#| label: shaprio_wilks_entropy_symbiont_potential

kw_entropy |>
  group_by(symbiont.potential) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

```{r}
#| label: shaprio_wilks_entropy_bleaching

kw_entropy |>
  filter(bleaching != "Pale") |>
  group_by(bleaching) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

```{r}
#| label: shaprio_wilks_entropy_location

kw_entropy |>
  group_by(location) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

```{r}
#| label: shaprio_wilks_entropy_Scleractinia_family

kw_entropy |>
  filter(host_family != "Plerogyridae", host_order == "Scleractinia") |>
  group_by(host_family) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

```{r}
#| label: shaprio_wilks_entropy_Scleractinia_symbiont_potential

kw_entropy |>
  filter(symbiont.potential != "Aposymbiotic", host_order == "Scleractinia") |>
  group_by(symbiont.potential) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

```{r}
#| label: shaprio_wilks_entropy_Scleractinia_bleaching

kw_entropy |>
  filter(!is.na(bleaching), host_order == "Scleractinia") |>
  group_by(bleaching) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

```{r}
#| label: shaprio_wilks_entropy_Scleractinia_location

kw_entropy |>
  filter(host_order == "Scleractinia") |>
  group_by(location) |>
  summarise(
    W = shapiro.test(MetabolicEntropy)$statistic,
    p_value = shapiro.test(MetabolicEntropy)$p.value
  )
```

## Levene **for Homoscedasticity** 

### Richness

```{r}
#| label: levene_richness_phylum

leveneTest(MetabolomicRichness ~ host_phylum, data = kw_richness)

```
